---
title: "Python troubleshooting"
author: "Charly Billaud, Junhong Yu"
date: "2024-12-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python troubleshooting}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

VertexWiseR uses Python functions from various toolboxes that can flexibly be called and executed in R using the *reticulate* package v1.37.0 [@ushey_reticulate_2023]. We explain in this article how to solve a number of issues that may arise from the interaction with Python.

## 1. About the message: "Would you like to create a default Python environment for the reticulate package? (Yes/no/cancel)"

As explained in the home page, this is a pop-up/prompt from the reticulate package. Choosing 'Yes' will let reticulate install Python in a virtual environment. This is entirely optional. Simply choose/click No/Cancel to ignore.

The [reticulate documentation](https://rstudio.github.io/reticulate/articles/python_packages.html) explains:

>When installing Python packages it’s best practice to isolate >them within a Python environment (a named Python installation >that exists for a specific project or purpose). This provides a >measure of isolation, so that updating a Python package for one >project doesn’t impact other projects. The risk for package >incompatibilities is significantly higher with Python packages >than it is with R packages, because unlike CRAN, PyPI does not >enforce, or even check, if the current versions of packages >currently available are compatible.

Note that VertexWiseR will install stable specific versions of Python itself and of each package it needs to work properly. If you believe you will be using the Miniconda/Python installation for more than VertexWiseR and will need to update them, it may be safer to install your specific Miniconda/Python environment in a path of your choice (as allowed by VWRfirstrun()), or to use a reticulate isolated virtual environment. More details on why use virtual environments can be found in the [reticulate documentation](https://rstudio.github.io/reticulate/articles/python_packages.html).

## 2. How to make VWRfirstrun() ignore the system's Python installation

VertexWiseR will automatically select a previous Python installation it finds via reticulate, but you may want to prevent this and run VertexWiseR on a new Python installation of its own. This section explains how to make `VWRfirstrun()` ignore a previous Python/Miniconda installation in order to install a new one (which will *not* affect the system's Python libraries).

Firstly, the Python installation path may already be preloaded in the global environment, via files such as `.Renviron`, `.Rprofile`, `.bashrc` etc. If that is the case they should be edited to remove the path of a previous Python installation, because reticulate will read the paths contained in these files.

Secondly, if the Python paths are not specified in the global environment, reticulate will then search for a Python installation in its default locations. This can be avoided with the method below.

To make `VWRfirstrun()` ignore a previous Python installation, you can set the `RETICULATE_PYTHON` variable as `NA` (in a new R session with a clean workspace):

```{r, eval=FALSE}
Sys.setenv(RETICULATE_PYTHON=NA)
VWRfirstrun()
```

With the above lines, reticulate will respond in the same way it would if Python was not previously installed at all. Therefore, `VWRfirstrun()` will let users install a new Miniconda or Python environment. After that, `VWRfirstrun()` will use that newly installed Python environment by default and you should not need to do this again.

## 3. How to make VWRfirstrun() use a previously installed Python

This section explains how you can choose a different Python other than the one automatically detected if they wish to.

### a. For a new VertexWiseR set-up

Reticulate allows you to predefine the Python environment to be used by entering its Python's directory path into the [following functions](https://rstudio.github.io/reticulate/reference/use_python.html):

```{r, eval=FALSE}
reticulate::use_python(python)
reticulate::use_virtualenv(virtualenv)
reticulate::use_condaenv(condaenv)
reticulate::use_miniconda(condaenv)
```

These functions must be executed before `VWRfirstrun()` is run in the R session. It should then automatically assume that a version of Python is correctly installed and only ask for other packages to be downloaded/installed if they are missing. To make sure the correct Python installation is being used, users can run:

```{r, eval=FALSE}
reticulate::py_config()
```

### b. If the VWRfirstrun() installations were already done before

If Python or Miniconda was already installed using `VWRfirstrun()`, then you will not be able to modify the Python path in the above manner. This is because `VWRfirstrun()` saves the path to the user-selected default or custom installations in order to access them again later. Every time `VWRfirstrun()` is run, it will read and use the saved paths.

The paths are saved in a `.Renviron` file, in the standard path generated via:

```{r, eval=FALSE}
tools::R_user_dir(package='VertexWiseR')
```

Within the `.Renviron` file, the environment variables will be written in this manner:

```{r, eval=FALSE}
#RETICULATE_MINICONDA_PATH="C:/path_to_miniconda/"
#RETICULATE_PYTHON_FALLBACK="C:/path_to_miniconda/"
#RETICULATE_PYTHON="C:/path_to_miniconda/python.exe"
```

You may choose to edit `.Renviron` directly, by replacing the paths with the new Python installation you would like to use. The `RETICULATE_MINICONDA_PATH` may also be safely removed if a non-conda Python library is to be used instead. Alternatively, users may choose to completely remove the `.Renviron` and define the path as described in section 3a.

## References:
