---
title: "Overlaying plots and transparent thresholding"
author: "Charly H. A. Billaud"
date: "2025-07-16"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overlaying plots and transparent thresholding}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(VertexWiseR)
```

## Transparent thresholding

This article lays out a number of functions users can use to overlay plots produced with surf_plot() in VertexWiseR. We show how we can use them to apply "transparent thresholding", which consists in reporting spatial neuroimaging results at a lower, non-significant subthreshold, to contextualise significant findings and improve broader interpretation across studies [@taylor2025go].

The demo data (218 MB) required to run the package demos can be downloaded from the package's github repository with the following function:

```{r, eval=TRUE}
#This will save the demo_data directory in a temporary directory (tempdir(), but you can change it to your own path)
demodata=VertexWiseR:::dl_demo(path=tempdir(), quiet=TRUE)
```

## Running a vertex-wise linear model

In this example, we use the Spreng dataset cortical thickness and behavioural data as in [Example 1](https://cogbrainhealthlab.github.io/VertexWiseR/articles/VertexWiseR_Example_1.html) and the [home  example](https://cogbrainhealthlab.github.io/VertexWiseR/index.html) , selecting young adults:

```{r, eval=TRUE}
SPRENG_CTv = readRDS(file=paste0(demodata,"/SPRENG_CTv_site1.rds"))
dat_beh = readRDS(paste0(demodata,'/SPRENG_behdata_site1.rds'))
#only young adults
SPRENG_CTv = SPRENG_CTv[dat_beh$agegroup=='Y',]
dat_beh = dat_beh[dat_beh$agegroup=='Y',]
#smoothing at 10 FMHW
SPRENG_CTv_smoothed = smooth_surf(SPRENG_CTv, 10, VWR_check = T)
```

We test the effect of sex on cortical thickness in the sample:

```{r, eval=TRUE}
#preparing model data
predictors=dat_beh[,c("site","age","sex")]
#running the model
results=RFT_vertex_analysis(model=predictors, 
                            contrast=predictors$sex, 
                            surf_data = SPRENG_CTv_smoothed,
                            p = 0.05 
                            )
results$cluster_level_results
```

According to these results, since the female sex is coded as 0 and males as 1, the regions colored in red are thicker in males.

## Plotting significant and subthreshold outcome

Both RFT_vertex_analysis() and TFCE_threshold() return thresholded and unthresholded t-maps.

Thresholded (significant) t-map:

```{r, eval=TRUE, results = 'hide'}
plot_surf(surf_data = results$thresholded_tstat_map,
          filename = paste0(tempdir(),"/sexdiff_thresh.png"),
          show.plot.window=FALSE)
```
```{r, echo=FALSE, fig.align="center", out.width="100%", fig.alt="Significant clusters after RFT correction"}
knitr::include_graphics(paste0(tempdir(),'/sexdiff_thresh.png'))
```

Unthresholded t-map:

```{r, eval=TRUE, results = 'hide'}
plot_surf(surf_data = results$tstat_map,
          filename = paste0(tempdir(),"/sexdiff_nothresh.png"),
          show.plot.window=FALSE)
```
```{r, eval=TRUE, echo=FALSE, fig.align="center", out.width="100%", fig.alt="Unthresholded t-statistic map"}
knitr::include_graphics(paste0(tempdir(),'/sexdiff_nothresh.png'))
```

## Overlaying plots together

To apply transparency thresholding, and letting information about subthreshold effects present in the plot, one may want to "merge" both maps together. 

plot_overlay() is a function that will allow you to plot a surface on top of another. To work in this context, we plot the thresholded t-stat map (surf_data_2) on top of the unthresholded one (surf_data_1), but we reduce the alpha opacity of the background surface to highlight the surface on top:

```{r, eval=TRUE, results = 'hide'}
plot_overlay_surf(surf_data_1=results$tstat_map,
                  surf_data_2=results$thresholded_tstat_map, 
                  cmap_1='RdBu_r', cmap_2='RdBu_r',
                  colorbar_1=FALSE, colorbar_2=TRUE,
                  alpha_1=0.3, alpha_2=1,
                  limits_2='same',
                  filename=paste0(tempdir(),'/overlay_plot.png'), 
                  title="Significant effect\nof sex",
                  show.plot.window=FALSE)
```
```{r, eval=TRUE, echo=FALSE, fig.align="center", out.width="100%", fig.alt="Significant vertices on top of transparent t map"}
knitr::include_graphics(paste0(tempdir(),'/overlay_plot.png'))
```
By default, both color bars are plotted but either can be set to FALSE (removed), which was done for the faded background colours here.

The function also give freedom to have two separate color maps, in order to highlight the significant t-values even more:

```{r, eval=TRUE, results = 'hide'}
plot_overlay_surf(surf_data_1=results$tstat_map,
                  surf_data_2=results$thresholded_tstat_map, 
                  cmap_1='RdBu_r', cmap_2='Reds',
                  alpha_1=1, alpha_2=1,
                  limits_2='same',
                  filename=paste0(tempdir(),'/overlay_plot_twocmaps.png'), 
                  title="Significant effect\nof sex",
                  show.plot.window=FALSE)
```
```{r, echo=FALSE, fig.align="center", out.width="100%", fig.alt="Significant vertices on top of transparent t map, with different color maps"}
knitr::include_graphics(paste0(tempdir(),'/overlay_plot_twocmaps.png'))
```

A similar procedure can be used if one wants the range of values to be independent (the default if the 'same' option is not provided):

```{r, eval=TRUE, results = 'hide'}
plot_overlay_surf(surf_data_1=results$tstat_map,
                  surf_data_2=results$thresholded_tstat_map, 
                  cmap_1='RdBu_r', cmap_2='hot_r',
                  alpha_1=1, alpha_2=1,
                  filename=paste0(tempdir(),'/overlay_plot_twolimits.png'), 
                  title="Significant effect\nof sex",
                  show.plot.window=FALSE)
```
```{r, echo=FALSE, fig.align="center", out.width="100%", fig.alt="Significant vertices on top of transparent t map, with different color maps and different ranges"}
knitr::include_graphics(paste0(tempdir(),'/overlay_plot_twolimits.png'))
```

## References: